---
title: "The Migrant Crisis Effect on the EU"
author: "Alina Kosiakova"
date: '2019-06-17'
output:
  pdf_document: default
  html_document:
    df_print: paged
    includes: null
editor_options:
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,collapse= TRUE, error = TRUE, warning=FALSE, message = FALSE, cache = TRUE)
```

```{r, echo=FALSE}
if(!require(eurostat)) install.packages("eurostat");
require(eurostat)
if(!require(dplyr)) install.packages ("dplyr");
require(dplyr)
if(!require(ggplot2)) install.packages ("ggplot2");
require (ggplot2)
```
INTRODUCTION

The migration crisis in the EU has led to a number of changes in the political, economic and social environments. According to the UN, arrivals in the year 2015 via the Mediterranean peaked at more than 1 million and polls have shown that immigrants have become voters number one concern and in a number of countries, such as France, Germany, Austria, Italy and Hungary, even the elections have been swayed to popularize populist and Eurosceptic parties. There are two vastly different opinions on the migration crisis – some people say it is a major threat that will destroy the EU, while others agree that it is a necessary thing to do to helps those in need and we will benefit massively from it in the future. In this essay, I will analyse the available data from neutral, non-political sources and conclude how the societies of the EU have really been affected by the migration crisis.

MAIN PART

I. What caused the migrant crisis and why did the EU become the target?

Officially we place the beginning of the migration crisis in 2015. In reality, that year is when Europe was hit by the storm of asylum seekers, but the trouble started brewing much earlier. The majority of the immigrants are people from Syria, Afghanistan and Somalia. There are multiple reasons why the immigration crisis started and why EU became the main destination:
1)	The Syrian civil war. According to the statistics compiled by the UN, the majority of asylum seekers come from Syria to Greece and Italy across the Mediterranean Sea. The people are afraid of war and run to protect themselves and their families from death – a very understandable fear and need. The reason why Europe is so attractive to them is because of legalities. The Syrian refugees who have immigrated to other Arab countries do not have the right to work, are not recognised as refugees and many of their children do not have access to basic education. Data provided by the UN shows that around 400000 Syrian children that are currently living in Turkey were rejected from the education system. Also, there has been a cut in UN funding which led to decreased financial help to Syrian families which makes the Middle East a highly unattractive place for people to stay.
2)	Encouragement from German Chancellor Angela Merkel. The EU laws say that the country to which an asylum seeker arrives first is the one held responsible for their application and resettling them, that is why countries such as Italy, Greece and Malta were heavily burdened by helping the refugees. A new wave of asylum seekers hit Europe when it was declared that if they manage to reach Germany, they could effectively apply for asylum there.
3)	Self-perpetuating element. The refugees that are lucky enough to reach Europe consequentially encourage their friends and families to try and immigrate to the EU as well. Also, the media’s intensive coverage on brief gaps in border control or impending caps on refugee numbers increase the number of people willing to try their luck as asylum seekers.
4)	Poverty and turmoil in sub-Saharan Africa. Many countries in that area of the continent are underdeveloped and poor. Before the instability in Libya, people would move there to find jobs and help their families, but now more people try to cross the Mediterranean. 

```{r, echo=FALSE}
migr_asyappctza <- get_eurostat ("migr_asyappctza",
                                 stringsAsFactors = FALSE)
```

```{r, echo=FALSE}
df <- migr_asyappctza %>%
  filter(geo %in% c("TOTAL"),
         time >= "2014-01-01",
         age == "TOTAL",
         sex == "T",
         asyl_app == "ASY_APP",
         citizen == "EXT_EU28",
       )%>%
  select(-c(1:6))

  
```         

```{r, echo=FALSE}
ggplot(df, aes(x=time, y=values, group=time))+
    geom_line(group=1)+
labs (title="Asylum applicants",
        subtitle = "Source: Eurostat (migr_asyappctza)",
        x="Year",
        y="Asylum applicants")
```
II. What kind of effect did the migrant crisis have on the economy of the EU?

The potential effect of the migration crisis on the economy is still very much up in the air. Many people are concerned that it might in fact weaken the economy by increasing unemployment, overloading the public budgets and straining the infrastructural capacity. Also, there is a fear that the refugee crisis will worsen the debt crisis. But is that really the case? We can divide the effects to short-term and long-term.
1)	Short-term. 
According to economist O. Reynolds, the short-term effects will be positive, but only slightly. When a refugee arrives to the host country, their health problems should be checked, children and adults must be educated to integrate them into the work labour markets, and also, financial support is required to help them before the adults are fully prepared to support their families. The cost to integrate a refugee into society varies from 8000 euros to 12000 euros per application (Kern, 2015). It means that the fiscal costs will come before the fiscal benefits for the host countries, but that will give an immediate positive boost to domestic demand. As stated in IMF, by the end of 2017 the GDP in Austria, Germany and Sweden – these countries have accommodated the most of refugees – had grown by 0.5%, 0.3% and 0.4%, respectively. According to The Economist, economic analysis demonstrates that the migrant increase will not increase the unemployment or decrease the wages of the already working native population. 

2)	Long-term. 
The results of the migration crisis are hard to predict. Not only because nothing major as this has not happened since World War II, but also because little is known about the education level and capabilities of the ones who have already immigrated to Europe. The process also heavily depends on how countries will be able to present the refugees to the work force and help them throughout the process of integration. In fact, the government only starts to receive fiscal benefits when the refugees enter the labour market. Also, the refugee crisis might be the solution to help an aging Europe with their declining birth-rate as the migrants will become net payers and contribute towards the increasing welfare expenditure on the pensioners. As of now, the current tendencies of gaining demographics will slow down the economic growth of Central Europe and Baltic nations. By welcoming the refugees, many of whom are young and skilled, may prove to be a lifeline for these economies because it would increase the number of people employed and contributing to the welfare system. Moreover, local businesses have reported that they received an increase in turnover and profits attributed to purchases made by the support staff, refugee customers and those engaged in public works. Refugees also tend to open their own small businesses according to information from Austria, Bulgaria and Sweden. The only problem remains – it is hard for businesses to employ asylum seekers and refugees due to their unclear residence status, language skills, level of qualifications, professional training and support needs, the recognition of diplomas, the lack of affordable housing and transport, xenophobia and bureaucratic hurdles for employers.


III. How did the migrant crisis change the political stances and the societies of the EU?

The biggest challenge in fact is the political climate of Europe. The migrant crisis consequentially led to the split between the older EU members such a Germany and the newer Central and Eastern European members. For example, they are divided on the quota system for refugee allocation to different EU counties. The debate among Member States had clearly demonstrated our differences not only in terms of fulfilling the EU obligations but also in the understanding of the value of human rights protection. Based on the scenarios that the European Commission has created (2017), it seems almost certain that their adoption will soon lead to the establishment of two different sides of the European Union. One of them will be the core, including Germany and hopefully France as leaders, and they would work to deepen their integration. Consequently, the so called “core” would diminish the role of countries which do not follow European obligations and fundamental rules. This poses a great threat to those countries, as this group includes the new Member States that benefit directly from the European integration process in the area of the Cohesion Policy, free movement of goods, services or capital. In the coming years they may become just a gateway to the European Union. Should that occur, the disintegration process will start, as for a Member State to become simply an entryway to the common market would be tantamount to EU’s disintegration.
The people have drastically different views on the migration crisis. According to Barysch, 2016, in a poll conducted in 2015, 60-80% of people in France, Italy, Spain and the UK claim they are displeased with their government’s immigration policies. Another poll in 2016 showed that 61% in France, 71% in Greece and 48% in the UK have feelings that their country’s membership in the EU is unfavourable in the current situation (Arnett 2016). These increasingly high numbers clearly indicate that the European citizens support the rise of right-wing, anti-EU and xenophobic political parties in many EU countries. For example, the far-right Conservative People’s Party of Estonia has almost 18% of votes in 2018, making in the third-largest part and Germany’s far-right party Alternative for Germany (AfG) entered the federal parliament for the first time with 12,6% of the votes and they became the biggest opposition party in the country. That explains the reasons why the wave of xenophobia has covered Europe. Never before have the European citizens been so sceptical about the European Union and its leaders. Hostility and mistrust have increased even in the most tolerant communities. Reception centres have met resistance in Denmark, Slovakia and 
Sweden and xenophobia and racism have become increasingly socially acceptable over social media.

CONCLUSION
The migrant crisis has become and will be one of the main problems of the European Union for quite some time. The findings of economist and finance analysts have shown that the refugee crisis might actually benefit Europe as a whole, not only increasing our turnover, but also helping an aging Europe. There will be challenges to integrate those people into society and help them find their way in the labour market, but the long-term benefits clearly outweigh the drawbacks. When it comes to the political consequences, it is clear that the EU is far from being integrated and the different Member States have drastically opposite opinions on the matter which rises from the general unsatisfaction of the general public who have started to support right-wing and sometimes even racist political parties, creating a more dangerous place for asylum seekers in Europe. It is clear that the EU will keep supporting the migrant crisis if the leaders of the union (Germany and France) have pro-migrant leaders. And only time will tell what happens next.

```{r}

```

@shubhampoddar2016
@europeanunionagencyforfundamentalrights2018
@aleksandraborowicz2017
@annehermmichelpoulain2018
@oliverreynolds2017
@elizabethcollettcamillelecoz



## References





